\section{Application Binary Interface}

\subsection{Introduction}

The LVX core Application Binary Interface (ABI) is a set of conventions
designed for inter-operability between the different code-generation tools and
libraries.

This ABI is little-endian only for bytes and bits.
Dynamic linking assumes that position-independent code and data are accessed
through pointers materialized in the called function.

The 32-bit addressing code generation uses a convention called ILP32, in which
integers, long integers, and pointers are 32-bit quantities. The 64-bit
addressing code generation uses the LP64 convention, in which integers are
32-bit quantities, while long integers and pointers are 64-bit quantities. In
both cases, \texttt{char} is a signed type and \texttt{long long} is a 64-bit
integer.

\subsection{Memory Addressing Model}

A program that executes on the LVX core views the addressable memory
as a collection of \emph{segments}. A segment is an area of memory that has
specific uses within an application and has no fixed address relationship to any
other segment. Thus, relative distances between any two items belonging to the
same segment are constant once the program has been linked, but the distance
between two items in different segments is not fixed. The program view of the
addressable memory as a collection of segments is independent from the
definition of address ranges with associated attributes in the MMU TLB entries.
It is expected, however, that the MMU be configured to enforce the correct
attributes of the segments once their addresses are resolved by the linker or
the loader.  The following memory segments are defined: \begin{description}

\item[Text] Contains the executable code, and exception tables.

\item[Data] Contains static storage variables with initialization values.

\item[Heap] Contains storage managed by dynamic memory allocation.

\item[Stack] Contains the stack frames of the executing functions.

\item[Device] Memory cells whose access triggers side-effects on hardware.

\end{description}

The memory segments except the device segment are populated by the linker or the
loader by concatenating the contributions from the program units that compose
the executable program. These contributions are contained into \emph{section}s,
whose mapping of to segments is as follows:

\begin{center}
\begin{tabular}{|l|l|l|} \hline
Section & Segment & Contents \\ \hline
\texttt{.text} & Text & Executable code \\
\texttt{.data} & Data & Initialized static data \\
\texttt{.bss} & Data & Cleared on start static data \\
\texttt{.rodata} & Data & Read-only data \\
\texttt{.tdata} & Data & Initialized thread-local data \\
\texttt{.tbss} & Data & Cleared on start thread-local data \\
\hline \end{tabular}
\end{center}


\subsection{Data Layout Rules}

\subsubsection{Native C Types}

Native types of the C language are \emph{naturally aligned}, that is, their
address must be a multiple of their size as computed by the \texttt{sizeof}
operator in C.

While the natural alignment of native types is not required by the LVX
architecture, it promotes more efficient code at little cost in storage.

\subsubsection{C Aggregate Types}

C aggregate types comprise arrays, complex types, GNU C vector types,
\verb|struct| types, and \verb|union| types: \begin{itemize}

\item An array is aligned on the same boundary as its element type.

\item Complex types and GNU C vector types inherit the alignment or their
    elements.  However, for performance reasons, they may be aligned on a
    multiple of their size in data sections and stack frames.

\item C \verb|struct| and \verb|union| types inherit the alignment of their most
    strictly aligned member.

\item The size of any aggregate type is always a multiple of its alignment.

\item Each aggregate type member is assigned to the lowest available offset with
    the appropriate alignment. This may require internal padding, depending on
    the previous member.

\item An aggregate size is increased, if necessary, to make it a multiple of the
    alignment. This may require tail padding, depending on the last member type.

\item There are no minimal size requirements for a \verb|struct| or \verb|union| type.

\end{itemize} The content of any padding is undefined.

\subsubsection{Bit-Field Layout}

C \verb|struct| and \verb|union| types may have bit-fields that define integral
objects with a specified number of bits.  The LVX run-time architecture
does not specify the signedness of bit-field types. This is specified by the
ISO~C99 language standard for all bit-field types except for plain \verb|int|.
The signedness of plain int bit-fields should be specified by the compiler
implementation.  Bit-fields obey the same size and alignment rules as other
\verb|struct| and \verb|union| members, with the following additions:
\begin{itemize}

\item Bit-fields are allocated least significant bit to most significant bit
    (from right to left).

\item A named bit-field must entirely reside in a storage unit appropriate for its
    declared type. For example, a bit-field of type short must never cross a
    half-word boundary. (This bit-field alignment rule is inherited from the
    historic PCC compiler.)

\item The type of unnamed bit-fields does not affect the alignment of the enclosing
    \verb|struct| or \verb|union| type.  Zero-length unnamed bit-fields force the
    alignment of subsequent members to the boundary corresponding to the size of
    the bit-field type.

\end{itemize}

\subsection{Stack Frame Organization}

\subsubsection{Stack Frame Addressing}

A function stack frame is an area of memory in the stack segment which is
dedicated to the activation of a called function.  The stack grows downwards,
and is delimited by the value of the stack pointer (SP) register that points to
its lowest address. An application may not write to memory below the stack
pointer.  Any memory below the stack pointer may be destroyed at any time, for
instance by interrupt handlers.  

The stack pointer must always be aligned on a 32-byte boundary.  This implies
that all stack frames must be a multiple of 32 bytes in size. When composing a
function stack frame, small object sizes and alignment are rounded up to 8
bytes. Objects that have stricter alignment than 8 bytes are assigned a slot
that ensures this stricter alignment, typically by allocating more stack slots
at compile time then rounding the address at execution time.

The frame grows downwards, meaning that local variable slots are addressed by a
negative offset from a virtual frame pointer. The first local variable slot is
thus at offset -8 from the virtual frame pointer.  The virtual frame pointer is
an abstraction and is different from the address maintained into the frame
pointer (FP) register, also known as hard frame pointer.  The FP register points
to the first address above the Dynamic Area region, described next.

\subsubsection{Stack Frame Regions}
\label{sss:stack-frame-region}

\begin{figure}
{\small
\begin{verbatim}
                            ~                ~
                            |  ............  |
                            |                |    ^
                            |                |    |
                            | Incoming       |    | Caller frame
        [256-bits aligned]  | Arguments      | <--/ <- incoming $sp
                            +----------------+
                            | Anonymous      |
                            | Arguments      |
                            |                |
                            +----------------+
                            |                |
                            | padding1       |
                            |                |
                            +----------------+
        [256-bits aligned]  | [Static chain] | <--- Virtual FP
                            +----------------+
                            | Local          |
                            | Variables      |
                            |                |
                            +----------------+
                            |                |
                            | padding2       |
                            |                |
                            +----------------+
                            |                |
                            | Register       |
                            | Save           |
                            |                |
                            | $ra            | (if frame_pointer_needed)
        [64-bits aligned]   | caller FP      | (<- $fp if frame_pointer_needed)
                            +----------------+
                            |                |
                            | [Dynamic Area] |
                            |                |
                            +----------------+
                            |                |
                            | padding3       |
                            |                |
                            +----------------+
                            |                |
                            | Outgoing       |
                            | Arguments      |
       [256-bits aligned]   |                | <- $sp
                            +----------------+
\end{verbatim}
}
\caption{Stack frame regions.} \label{fig:stack-frame}
\end{figure}

A function stack frame comprises up to six regions, listed from the
highest address to the lowest address in the stack frame
(Figure~\ref{fig:stack-frame}): \begin{description}

\item[Anonymous Arguments] This region contains space for storing the
    incoming anonymous arguments passed in registers in case the function is
    variadic. Anonymous arguments, also known as unnamed arguments, are those
    passed at position starting at ellipsis '\texttt{...}' in the prototype
    of a variadic function.

\item[Local Variables] A function may store local variables and
    compiler temporaries in this region. In case a static chain pointer is
    required, it uses the first local variable slot. The static chain pointer is
    used in languages that support nested function definitions, in order to
    access the local variables of the enclosing functions. Nested functions are
    supported as an extension in GNU C.

\item[Registers Save] A function may store callee-saved (preserved)
    registers in this region before reusing them. A function may also store
    caller-saved (scratch) registers in this region before calling a function.

\item[Frame Marker] This region contains information required for
    unwinding the stack at run-time, and is pointed to by the FP register. It
    contains a pointer to the previous frame marker (the caller FP), and the
    return address of the function. The pointer is NULL if there is no previous
    frame marker.

\item[Dynamic Area] This region is variable-sized and created in case the
    function calls \verb|alloca()| or declares ISO~C99 variable length arrays.
    When the Dynamic Area exists, a Frame Marker region is required. Otherwise,
    the Frame Marker region is optional.

\item[Outgoing Arguments] Successive arguments are passed in argument slots,
    with the first twelve ones located in registers R0--R11, and the other on
    the stack in the Outgoing Arguments region. The Outgoing Argument region on
    the stack is sized as the maximum required by all functions called from the
    current one.  Stack space is not allocated for the arguments that are passed
    in registers, whether named or anonymous.

\end{description}

\subsection{Nested Functions Implementation}

As mentionned in \ref{sss:stack-frame-region}, a stack frame may need a static
chain slot which uses the first local variable slot. This slots stores a
pointer to the location of the local variables of its caller (whose activation
record is stored higher in the stack).

Let's recall that the situation is pretty different for the caller (the
function which has a nested function) and the callee (the nested function
itself), even though a function can be both at the same time if there is a
nested function in a nested function.

The caller has to give away a pointer pointing on its local variables so that
the callee can store it in the its static chain slot for later use. Currently
this is done by using GPR \texttt{\$r31} (a callee saved register used as a
scratch register). When calling a nested function, the calling function:
\begin{enumerate}
    \item Store its static chain in \texttt{\$r31},
    \item Call the nested function
    \item The nested function stores the address of the static chain in a
          special slot in its frame. This is done in the prologue of the function, once
          \texttt{\$r31} has been saved it can be used again, saving \texttt{\$r31} is
          espacially useful when the depth of the nest of nested function is greater than
          one (ie, there is at least one nested function in a nested function).
\end{enumerate}

In case there is a need to pass the address of a nested function to another
function (whose stack frame will be below the function which performs the call
(the general case is not covered)), the calling function sets up a trampoline
of size (12 * 2 + 4 = 28) on the stack composed of:

\begin{verbatim}
    0:  make $r31 = ptr to static chain of caller
    c:  make $r30 = ptr to the nested function
   10:  igoto $r30 (Note that icall would destroy $ra.)
\end{verbatim}

Once this is done, the calling function gives the address of the trampoline
instead of the real address of the nested function. This way, when it is called
it first setup the static chain.

\subsubsection{Regarding cache alignement and cache invalidation}

The macro \texttt{TRAMPOLINE\_ALIGNEMENT = 32}, should guarantee by
construction that the trampoline is correctly aligned on the stack and the
three instructions (each in a separate bundle) should not overlap two cache
lines such that the a call to \texttt{i1invals (static\_chain\_address)}
clears the cache line where those $3$ instructions are.

\subsection{Function Arguments and Result}

Arguments are passed in a combination of general registers and memory, as
described below. Arguments are first allocated on a conceptual argument list,
that begins at relative address 0, and are allocated towards increasing
addresses. Each storage unit in the argument list, called argument slot, has a
8-byte size and alignment. Argument slots are referred to as slot~0, slot~1,
slot~2, etc.  Arguments are aligned to 8-byte boundary and padded within the
argument list according to the following rules: \begin{itemize}

\item  Integral scalar arguments smaller than 64 bits are padded on the left to
    a total width of 64 bits (that is, the value is placed in the
    least-significant bits of the argument slot), and placed in the next
    available argument slot.

\item Half-precision floating-point arguments are not promoted to
    single-precision floating-point. They are passed directly with 48-bit of
    padding.

\item Single-precision complex arguments are passed as a single argument.  The
    real part is passed in the 32 least significant bits and the imaginary part
    in the 32 most significant bits.

\item Double-precision complex arguments are passed as a pair of
    double-precision floating-point scalar arguments. The real part is passed
    first and the imaginary part second.

\item In C++, aggregates that are not trivially copyable classes are always
    passed by reference. A trivially copyable class has no virtual functions or
    virtual base classes.

\item Aggregates up to 8 bytes in size, when passed by value, are padded to a
    total width of 64 bits and placed in the next available argument slot. If
    shorter than 8 bytes, the aggregate begins at the lowest-numbered byte in
    the storage unit, so the padding is on the highest-numbered bytes.

\item Aggregates and data types larger than 8 bytes and up to 32 bytes are
    passed by value. They are padded to a multiple of 64 bits, aligned to an
    8-byte boundary, and placed in as many argument slots as needed. Any
    padding required is on the highest-numbered bytes of the final argument slot.

\item Aggregates and data types of size larger than 32 bytes are passed by
    reference, with a pointer using a full argument slot in place of the
    argument itself. It this the caller's responsibility to allocate a memory
    buffer for a copy of the original argument, copy into it, and pass the
    pointer to this copy in the corresponding argument slot.

\item As an extension, GNU C allows variable-length arrays as members of
    \verb|struct| or \verb|union| types. Such aggregates have a variable size
    and are always passed by reference.

\item If an argument straddles the boundary between slot~11 and slot~12, the
    part that lies within the first twelve slots is passed in general registers,
    and the remainder is passed in the Outgoing Arguments region.

\end{itemize} The content of any padding is undefined.

The first 12 arguments slots of the argument list are allocated into the
general-purpose registers R0--R11. If more than 12 slots are required by the
argument list, these are allocated in the Outgoing Arguments region with
increasing addresses, starting by the 8-byte stack slot pointed to by the
current SP (Figure~\ref{fig:stack-frame}). So slot~12 is placed at location SP,
slot~13 at SP+8, and so on.  When allocating arguments to slots, the only
alignment enforced is 8 bytes.  This is possible because the LVX
architecture does not require any memory alignment for correct behavior. As a
result, some multi-register arguments may end up in mis-aligned register
tuples. It is the callee's responsibility to move such arguments into
well-aligned register tuples before using them as operands of wide load-store
and SIMD instructions. Likewise, it is the caller's responsibility to move such
arguments from well-aligned register tuples.

The rules above support variable-argument list functions in both the K\&R~C and
the ISO~C99 languages. Since the LVX core does not specialize registers,
there is no issue whether the presence or absence of a function prototype
changes the way arguments are passed.  Thus, a variadic function may assume that
the anonymous arguments that lie within the first twelve argument slots can all
be found in the argument passing general registers, R0--R11. It may then store
these registers to memory, providing up to 96 bytes of Anonymous Argument region
on top of its own stack frame for R0--R11, as necessary
(Figure~\ref{fig:stack-frame}). This arrangement places all the anonymous
arguments in one contiguous block of memory.

Return values up to 32 bytes (4$\times$8) bytes are returned directly in
registers, according to the same rules as if they were to be passed as the first
parameter of an argument list. Return values larger than 32 bytes or of dynamic
size are returned in a memory buffer allocated by the caller. In this case, a
pointer to the buffer is passed to the called function in R15.


\subsection{Function Linkage}

\subsubsection{Register Usage Conventions}

Depending on the code generation model, the LVX core registers must
obey conventions for proper function linkage. The function linkage
conventions specify agreement between the caller and the callee on the following
register usage: \begin{description}
\item[reserved] All registers that are not allocatable (such as SP) must be reserved.
\item[argument] List of function call argument registers.
\item[result] List of function call result registers.
\item[struct] The result pointer for functions that return a \verb|struct| or \verb|union| type.
\item[caller] List of caller-saved registers, also known as scratch registers.
\item[callee] List of callee-saved registers, also known as preserved registers.
\item[program] The program counter register.
%\item[handler] List of C++ exception handler result registers.
%\item[veneer] List of intra-function-call temporary registers, destroyed by CALLs.
\item[return] The return address register.
\item[stack] The stack pointer (SP) register.
%\item[static] The static link pointer register (caller frame).
\item[frame] The frame pointer (FP) register.
%\item[global] Global data or Global Offset Table (GOT) pointer.
\item[local] Local data or Thread Local Storage (TLS) pointer.
%\item[wired] Wired register, read and write have no effects.
%\item[zero] Always read as zero or false register.
%\item[one] Always read as one or true register.
\end{description}
Although not listed above, the coprocessor registers are considered caller-saved
(scratch). It is recommended to use the coprocessor registers only in leaf
functions, as saving and restoring them implies a significant run-time cost.

When compiling, static storage data are accessed by building addresses in code.
Register usage conventions for the LVX code generation mode are defined in
Table~\ref{table:regular}. When compiling in PIC (Position-Independent Code)
mode, static storage data are accessed relative to a Global Pointer (GP) that is
set depending on the load module the function belongs to. The LVX PIC code
generation assumes the callee sets the GP.

\begin{table}
\begin{center}
\input{Convention-regular}
\end{center}
\caption{Register usage conventions.}
\label{table:regular}
\end{table}

\subsubsection{Function Call Sequence}

Direct and indirect function calls follow the same sequence before the call and
after the call. The function calling sequence is as follows: \begin{description}

\item[Call preparation] Values in caller-saved (scratch) registers that must be
kept live across the call must be saved. They can be saved by either copying
them into preserved registers or by saving them into the function stack frame.

\item[Call instruction] Use CALL or ICALL for PC-relative direct calls and
indirect calls respectively. These instructions write the value of the PC of
the next bundle in register RA, then change the PC to the target address of the
call.

For direct calls the PC-relative displacement to the target is computed at
link time.  Compilers may assume that the standard displacement field in the
call instruction is sufficiently wide to reach the target of the call. If the
displacement is too large, the linker must supply a branch stub at some
convenient point in the code; compilers must guarantee the existence of such a
point by ensuring that code sections in the relocatable object files are no
larger than the maximum reach of the call instruction.

\item[Function entry] The prologue code in the target function is responsible
for allocating a frame on the memory stack, if necessary.  If it is a
non-leaf function, it must save the RA register in the stack frame.  The
prologue must also save any callee-saved (preserved) registers that will be used
in this function.

\item[Function exit] The epilogue code is responsible for restoring the RA
register and any callee-saved (preserved) registers that were saved.  If a
memory stack frame was allocated, the epilogue code must de-allocate it.
Finally, the function exits by branching through the RA register with the RET
instruction.

\item[After the call] Any caller-saved (scratch) values should be restored.

\end{description}

A compiler can choose to make a tail call (including tail recursion) by simply
branching to the function being called. However, this appears to a debugger as
if the nested function was called directly from the caller of the original
function.  If the calling function has a stack frame, it must pop its frame off
the stack before making the call. Arguments should be passed in the usual
registers.  Tail calls can only be made when all parameters can be passed in
registers.


\subsection{System Call Interface}

From the ABI point of view, system calls are seen as regular function calls, so
calling conventions are assumed. The system call number is available to the
system call handler in the SN field of the PS register.


\subsection{Relocation List}

\begin{center}
\input{Relocation}
\end{center}


%\subsection{ELF Specifications}

%\subsection{DWARF Register Numbers}

%\input{DwarfId}


%\subsection{Compiler Intrinsic Functions}


